---
 title: "diffCircadian tutorial"
 auther: "Zhiguang Huo, Haocheng Ding (Department of Biostatistics, University of Florida)"
 date: "`r Sys.Date()`"
 output: 
   html_document:
    toc: true
    toc_depth: 2
    number_sections: true
 vignette: >
  %\VignetteIndexEntry{diffCircadian}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(stringsAsFactors = FALSE)
```

# Introduction

## Background
Circadian rhythmicity in transcriptomic profiles has been implied in many physiological processes, and disruption of circadian patterns has been founded to associate with several diseases. Likelihood-based methods (diffCircadian) are powerful in detecting circadian rhythmicity within one experimental condition and differential circadian patterns comparing two experimental conditions. 

## Statistical method
Denote $y$ as the expression value for a gene; 
$t$ as the circadian time; $C$ as the basal level (vertical shift of the sinusoidal wave baseline from 0); 
$A$ as the amplitude. 
$\omega$ is the frequency of the sinusoidal wave, where $\omega = \frac{2\pi}{Period}$. 
Without loss of generality, we set $period = 24$ hours to mimic the diurnal period. 
$\phi$ is the phase of the sinusoidal wave curve. Due to the periodicity of a sinusoidal wave, $\phi$ is not identifiable in the sense that $\phi = \phi + 24$. Therefore, we will restrict $\phi \in [-6, 18)$. $\phi$ is difficult to read from a sinusoidal wave, but a closely related quantify is the peak time $t_P$. The connection between $\phi$ and $t_P$ is that $\phi + t_P = 6 \pm 24N$, where $N$ is an arbitrary natural number.


### Circadian Rhythmicity Detection
We assume:

\begin{equation}
\label{eq:sin1}
y_i=A\sin(\omega(t_i+\phi))+C+\varepsilon_i, \qquad (1) 
\end{equation}

\noindent where $\varepsilon_i$ is the error term for subject $i$; 
we assume $\varepsilon_i$'s are identically and independently distributed (i.e., $iid$) and $\varepsilon_i \sim  \textit{N}(0, \sigma^2)$, 
where $\sigma$ is the noise level. 
To benchmark the goodness of sinusoidal wave fitting, 
we define the coefficient of determination $R^2=1-\cfrac{RSS}{TSS}$, 
where $RSS = \sum^n_{i=1} (y_i-\hat{y_i})^2$, 
$TSS = \sum^n_{i=1} (y_i-\bar{y})^2$,
$\hat{y_i} = \hat{A} \sin(\omega(t_i+ \hat{\phi})) + \hat{C}$,
$\bar{y} = \sum_i y_i / n$,
with $\hat{A}$, $\hat{\phi}$, and  $\hat{C}$ being the fitted value for $A$, $\phi$, and $C$ in equation (1), respectively.
$R^2$ ranges from 0 to 1, with 1 indicating perfect sinusoidal wave fitting, 
and 0 indicating no fitting at all.
Based on these assumptions, 
we derive procedures for testing circadian rhythmicity. 
For the ease of discussion, we re-write equation (1) as 

\begin{equation}
\label{eq:sin2}
y_i=E\sin(\omega t_i) + F\cos(\omega t_i) +C+\varepsilon_i, \qquad (2) 
\end{equation}
where $E = A \cos (\omega \phi)$, and $F = A \sin (\omega \phi)$.
Denote $\pmb{\beta}  = (E,F,C,\sigma^2)^\boldsymbol{\top}$,
the null hypothesis is $H_0: \pmb{\beta} = \pmb{\beta}_0 = (0, 0, \hat{C}_0,\hat{\sigma}^2_0)^\boldsymbol{\top}$,
where $\pmb{\beta}_0$ is the least square estimate of equation (2) under $H_0$. 
The null hypothesis indicates there does not exist a circadian rhythmicity.
On the other hand,
the alternative hypothesis is $H_A: \pmb{\beta} = \pmb{\beta}_A = (\hat{E}_A, \hat{F}_A, \hat{C}_A,\hat{\sigma}^2_A)^\boldsymbol{\top}$,
where $\pmb{\beta}_A$ is the least square estimate of equation (2) under $H_A$. 
The alternative hypothesis indicates there exists a circadian rhythmicity.

#### Likelihood Ratio Test 

Based on equation (2), the likelihood function of all $n$ samples is: 

\begin{equation*}
    L(E,F,C,\sigma^2) = \prod^n_{i=1} L_i(E,F,C,\sigma^2|y_i,t_i)= \prod^n_{i=1} \cfrac{1}{\sqrt{2\pi\sigma^2}}
    \exp\left( -\cfrac{ (y_i-E\sin(\omega t_i) - F\cos(\omega t_i)-C)^2}{2\sigma^2} \right)
\end{equation*}

\noindent The log-likelihood function is: 
\begin{equation*}
    l(E,F,C,\sigma^2) = \log L(E,F,C,\sigma^2) = 
    -\cfrac{n}{2} \log (2\pi\sigma^2)-\sum^{n}_{i=1}\cfrac{(y_i-E\sin(\omega t_i) - F\cos(\omega t_i) - C)^2}{2\sigma^2}
\end{equation*}

\noindent Under $H_0$, 
$\pmb{\beta} = \pmb{\beta}_0 = (0, 0, \hat{C}_0,\hat{\sigma}^2_0)^\boldsymbol{\top}$
and $l_0 = l(0, 0, \hat{C}_0,\hat{\sigma}^2_0)$;
under $H_A$, $\pmb{\beta} = \pmb{\beta}_A = (\hat{E}_A, \hat{F}_A, \hat{C}_A,\hat{\sigma}^2_A)^\boldsymbol{\top}$ and $l_A = l(\hat{E}_A, \hat{F}_A, \hat{C}_A,\hat{\sigma}^2_A)$.
The likelihood ratio test statistic is: $T^{LR} = -2(l_0-l_A)$. Since the degree of freedom is 2, under $H_0$, $T^{LR} \sim \chi^2_2$.

#### Wald Test
The Wald test statistic can be derived as
$T^{Wald} = (\pmb{\beta}_A-\pmb{\beta}_0)^\boldsymbol{\top} I(\pmb{\beta}_A)(\pmb{\beta}_A-\pmb{\beta}_0)$, where $I(\pmb{\beta}_A)$ is Fisher information matrix evaluated at $\pmb{\beta}_A$. Under $H_0$, $T^{Wald} \sim \chi^2_2$.

#### Finite Sample Wald/LR Tests
The finite sample Wald statistics ($T^{Wald}_{FN}$) and 
the finite sample likelihood ratio statistics ($T^{LR}_{FN}$) can be derived as the following:

\begin{equation}
   T^{Wald}_{FN} = \cfrac{n-k + 1}{nr} T^{Wald}
   \label{eq:Wald_FN}
\end{equation}

\begin{equation}
    T^{LR}_{FN} = \frac{n - k + 1}{r} \left(\exp(T^{LR}/n) - 1 \right) 
   \label{eq:LR_FN}
\end{equation}

\noindent Where $k = 4$ is total number of parameters and $r=2$ is number of parameters of interest. 
Under the null hypothesis, 
$T^{Wald}_{FN} \sim F(df_1, df_2)$, and $T^{LR}_{FN} \sim F(df_1, df_2)$,
where $df_1 = r$ and $df_2 = n - k + 1$.


#### F Test
\begin{equation*}
T^{F}=\cfrac{(TSS-RSS)/df_1}{RSS/df_2}\sim F(df_1,df_2), 
\end{equation*}

\noindent
where residual sum of squares $(RSS)=\sum^{n}_{i=1}(y_i-\hat{y_i})^2$, total sum of squares $(TSS)=\sum^{n}_{i=1}(y_i-\bar{y})^2$, $df_1=2$ and $df_2=n-3$. Under the null hypothesis, 
$T^{F} \sim F(df_1, df_2)$.


### Differential Circadian Detection

Denote $y_{1i}$ as the gene expression value of subject $i (1\le i \le n_1)$ in experimental condition 1, 
where $n_1$ is the total number of subjects; $t_{1i}$ is the circadian time for subject $i$.
$y_{2j}$ is the gene expression value of subject $j (1\le j \le n_2)$ in experimental condition 2, 
where $n_2$ is the total number of subjects; $t_{2j}$ is the circadian time for subject $j$.
Note that $y_{1i}$ and $y_{2j}$ are from the same gene, but under different experimental conditions.
We assume the following models:
\begin{equation}
  \label{eq:diffCircadian}
  \begin{aligned}
    y_{1i} &= A_1\sin(\omega(t_{1i}+\phi_1))+C_1+\varepsilon_{1i} \\
    y_{2j} &= A_2\sin(\omega(t_{2j}+\phi_2))+C_2+\varepsilon_{2j}
  \end{aligned}
\end{equation}
\noindent $\varepsilon_{1i} \sim \textit{N}(0,\sigma_1^2)$ is the error term for subject $i$ ($1\le i \le n_1$) for experimental condition 1 and 
$\varepsilon_{2j} \sim \textit{N}(0,\sigma_2^2)$ is the error term for subject $j$ ($1\le j \le n_2$) for experimental condition 2. 
These error terms are assumed to be $iid$.
$A_1$, $\phi_1$, $C_1$, and $\sigma_1^2$ are the amplitude, phase, basal level, and noise level for the experimental condition 1, and
$A_2$, $\phi_2$, $C_2$, and $\sigma_2^2$ are for experimental condition 2. 

#### Hypothesis Testing Framework 
Below we state the null hypothesis and the alternative hypothesis for testing these four categories of differential circadian patterns.


(1) Differential amplitude: $H_0: A_1 = A_2 = A_c$ \textit{v.s.} $H_A: A_1 \neq A_2$. 

(2) Differential phase: $H_0: \phi_1 = \phi_2 = \phi_c$ \textit{v.s.} $H_A: \phi_1 \neq \phi_2$.

(3) Differential basal level: $H_0: C_1 = C_2 = C_c$ \textit{v.s.} $H_A: C_1 \neq C_2$.

(4) Differential rhythmicity: $H_0: \sigma^2_1 = \sigma^2_2 = \sigma^2_c$ \textit{v.s.} $H_A: \sigma^2_1 \neq \sigma^2_2$.

#### Likelihood Ratio Test

The log-likelihood function for $n_1 + n_2$ samples in both experimental conditions is:
\begin{equation}
\begin{split}
    & ll (A_1, \phi_1, C_1, \sigma_1^2, A_2, \phi_2, C_2, \sigma_2^2) \\
    & =  \sum_{i=1}^{n_1} l_i (A_1, \phi_1, C_1, \sigma_1^2) + 
    \sum_{j=1}^{n_2} l_j (A_2, \phi_2, C_2, \sigma_2^2) \\
    & =  \left(-\cfrac{n_1}{2} \log \sigma_1^2-\sum^{n_1}_{i=1}\cfrac{(y_i-A_1\sin(\omega(t_i+\phi_1))-C_1)^2}{2\sigma_1^2}\right) \\
    & + \left(-\cfrac{n_2}{2} \log \sigma_2^2-\sum^{n_2}_{j=1}\cfrac{(y_j-A_2\sin(\omega(t_j+\phi_2))-C_2)^2}{2\sigma_2^2}\right)
\end{split} 
\end{equation}
The test statistic is: $D^{LR} = -2(ll_0-ll_A)$,  
where $ll_0$ is the log likelihood under $H_0$; and $ll_A$ is the log likelihood under $H_A$.
Here the null can be one of the following:
(a) $H_0^{(a)}$: $A_1 = A_2$ for differential amplitude; 
(b) $H_0^{(p)}$: $\phi_1 = \phi_2$ for differential phase; 
(c) $H_0^{(b)}$: $C_1 = C_2$ for differential basal level; 
(d) $H_0^{(r)}$: $\sigma^2_1 = \sigma^2_2$ for differential rhythmicity.
For all these null hypotheses, the degree of freedom is 1, and under $H_0$, $D^{LR} \sim \chi^2_1$.
For example when testing different amplitude, 
under $H_0^{(a)}$, $\hat{A_1}=\hat{A_2}=\hat{A_c}$, $ll_0 = ll(\hat{A_c}, \hat{\phi_1}, \hat{C_1},\hat{\sigma_1},\hat{A_c}, \hat{\phi_2},\hat{C_2},\hat{\sigma_2}|y)$; 
and under $H_A^{(a)}$, $\hat{A_1} \neq \hat{A_2}$, $ll_A = ll(\hat{A_1},\hat{\phi_1}, \hat{C_1},\hat{\sigma_1},\hat{A_2},\hat{\phi_2},\hat{C_2},\hat{\sigma_2}|y)$. 

#### Wald Test
Denote $\pmb{p} = (A_1, \phi_1, C_1, \sigma_1^2, A_2, \phi_2, C_2, \sigma_2^2)^\boldsymbol{\top}$.
$\pmb{p}_0$ is $\pmb{p}$ under $H_0$, 
where $H_0$ is one of the null hypotheses in section $\boldsymbol{1.2.2.1}$;
$\pmb{p}_1$ is $\pmb{p}$ under $H_A$, where there is no restriction on $\pmb{p}$.
Then the Wald test statistic is
$D^{Wald} = (\pmb{p}_1-\pmb{p}_0)^\boldsymbol{\top} I(\pmb{p}_1)(\pmb{p}_1-\pmb{p}_0)$.
Under $H_0$, $D^{Wald} \sim \chi^2_1$,
where 
$I(\pmb{p}_1)$ is Fisher information matrix evaluated at $\pmb{p}_1$.  


## About this tutorial
This is a tutorial for the usage of "diffCircadian" package. A real data example of aging on circadian patterns of gene expression in the human prefrontal cortex is used. The major contents of this tutorial includes: 
- How to prepare input for diffCircadian.
- Detection of circadian and differential circadian patterns.
- Visualizing circadian patterns' curves.

# About the package

## How to install the package

To install this package, start R (version "3.6" or higher) and enter: 

```{r, eval=FALSE}
library(devtools)
install_github("diffCircadian/diffCircadian") 
```

## How to cite the package

## Maintainer
Haocheng Ding (haochengding@ufl.edu)

## Description about the example data
Brain aging data were collected from University of Pittsburgh's Brain Tissue Donation Program. The final samples included 146 individuals whose time of death (TOD) could be precisely determined. The mean age at death was 50.7 years; 78% of the individuals were male, and the mean postmortem interval was 17.3 hours. The TODs were further adjusted as the Zeitgeber time (ZT), which considered factors including time zone, latitude, longitude, and altitude.
The ZT was used as the circadian time, which was comparable across all individuals. 33,297 gene probes were available in this microarray data, which was publicly available in GEO (GSE71620). After filtering 50% gene probes with lower mean expression level, 16,648 gene probes were kept. 


## Read in the example data
```{r}
library(diffCircadian) # include diffCircadian package
# load the data
data(BA11)
data(BA11old)
data(BA11young)
```


# Circadian Rhythmicity Detection

## Test all genes in BA11 data
```{r}
sig_index <- NULL # index of gene with p-value < 0.05
pvalue <- NULL
tod <- as.numeric(BA11[1,-1]) # Time of Death (TOD)
yy <- BA11[-1,]
yy <- yy[,-1] # get expression value matrix

for (i in 1:nrow(yy)){
  p <- LR_rhythmicity(tod,as.numeric(yy[i,]),FN=F)$pvalue
  pvalue <- c(pvalue,p)
}
```

## Plot top significant genes with circadian rhythmicity
```{r echo=F}
library(dplyr)
BA11_plot <- cbind(BA11[-1,],pvalue)
BA11_plot <- BA11_plot %>% arrange(pvalue)
BA11_plot_top6 <- BA11_plot[1:6,]
for (i in 1:nrow(BA11_plot_top6)){
  
  LR <- LR_rhythmicity(tod,as.numeric(BA11_plot_top6[i,2:293])) 
  amp <- LR$amp
  phase <- LR$phase
  offset <- LR$offset
  plot(tod,as.numeric(BA11_plot_top6[i,2:293]),pch=20,ylab="Expression Value",xlab = "Time of Death (TOD)",main = paste(BA11_plot_top6[i,1], "p-value = ",signif(BA11_plot_top6[i,294],3)),xlim=c(-10,20))
  f <- function(x){amp*sin(pi/12*(x+phase))+offset}
  curve(f,add=T,col="red",lwd=3)
}
                            
```

# Differential Circadian Pattern Detection

## Select genes with significant circadian pattern in both old and young groups

```{r}
TOD_old <- as.numeric(BA11old[1,-1])
TOD_young <- as.numeric(BA11young[1,-1])
index <- NULL
for (k in 2:nrow(BA11)){
  p_young <- LR_rhythmicity(TOD_young,as.numeric(BA11young[k,-1]))$pvalue
  p_old <- LR_rhythmicity(TOD_old,as.numeric(BA11old[k,-1]))$pvalue
  if(p_young < 0.05 & p_old < 0.05){
    index <- c(index,k)
  }
}
```

## Differential Amplitude

```{r}
for (i in 2:nrow(BA11)){
  p_amp <- LR_diff(TOD_young,as.numeric(BA11young[i,-1]),TOD_old,as.numeric(BA11old[i,-1]),type="amp")$pvalue
  if(p_amp < 0.05){
    index_amp <- i # index of first gene detected with differential amplitude.
    break
  }
}

LR_diff_amp <- LR_diff(TOD_young,as.numeric(BA11young[index_amp,-1]),TOD_old,as.numeric(BA11old[index_amp,-1]),type="amp") # perform LR differential amplitude test 


# plot result
plot(TOD_young,as.numeric(BA11young[index_amp,-1]),pch=20,ylab="Expression Value",xlab="Time of Death (TOD)", main="Age Young Group")
amp_y <- LR_diff_amp$amp_1
phase_y <- LR_diff_amp$phase_1
offset_y <- LR_diff_amp$offset_1

amp_o <- LR_diff_amp$amp_2
phase_o <- LR_diff_amp$phase_2
offset_o <- LR_diff_amp$offset_2

f_y <- function(x){amp_y*sin(pi/12*(x+phase_y))+offset_y}
curve(f_y,add=T,col="red",lwd=3)

```


```{r}
sessionInfo()
```